Comparing MCMC algorithms for Stochastic Volatility Model Bayesian Posteriors using Simulation Based Calibration

### 2023-10-16 manuscript/main.tex

\title{Comparing MCMC algorithms in Stochastic Volatility Models using Simulation Based Calibration}
\author{Benjamin Wee}
\date{October 2023}

@@

\begin{abstract}
    This research compares the computational methods used to estimate Bayesian stochastic volatility models. The MCMC approaches outlined in the stochastic volatility literature and Hamiltoninan Monte Carlo are assesssed in their ability to sample from the model's posterior distribution. Specifically, Simulation Based Calibration (SBC) is used to check whether these MCMC algorithms are returning efficient and calibrated posterior estimates. Key metrics of interest are the effective sample size to check the efficiency of the algorithm and tests of uniformity to assess the calibration of the posteriors. This will determine which method is better at estimating stochastic volatility models based on the efficiency and accuracy of the sampling strategy. Results reveal that Hamiltonian Monte Carlo provides more efficient and calibrated posterior estimates of the stochastic volatility model conditional on model paramterisation. 
\end{abstract}

@@

\section{Introduction}
    The stochastic volatility model is used in financial econometrics to model the behaviour of financial assets. They are typically expressed as non linear Gaussian state space models where the variance is treated as a latent random variable. Stochastic volatility is difficult to estimate since the likelihood is unavailable in closed form and there are at least as many variables as data points. Bayesian methods can be used to estimate these models which rely on Markov Chain Monte Carlo (MCMC) techniques to sample from the target joint posterior of these high dimensional parameter spaces.

    \citet{kim1998stochastic} proposed a sampling strategy for stochastic volatility through a combination of conjugate posterior distributions, a Kalman filter, and the Metropolis Hastings algorithm. Since then, there have been many advances in statistical computing and algorithm design. In particular, Hamiltonian Monte Carlo - a variant of the Metropolis Hastings algorithm - is an advancement in MCMC for efficient sampling of complex, high dimensional probabilistic models. The adoption of such new techniques have been made widely available through the development of various open source libraries such as the Stan programming language \citep{stan} and the PyMC library \citep{pymc2023}. Other algorithms have also improved the speed of estimating high dimensional models using approximations of the posterior such as integrated nested Laplace approximation \citep{rue2009approximate}.

    These conceptually and practically different techniques to estimate complex models did not exist 20 years ago (or perhaps more accurately, were not easily accessible or implemented 20 years ago). As the development of new algorithms rapidly increase, so does the need to develop new methods to assess their output. Developments in statistical workfow are required to test new algorithms as well as compare computational strategies used to estimate increasingly complex models.

    This research conducts a simulation study to analyse the calibration and efficiency of MCMC used to estimate stochastic volatility models. The first algorithm replicates Kim Sherphard and Chib's (KSC) Gaussian mixture approximation which is sampled using the aforementioned Kalman Filter and Metropolis Hastings algorithm. The second algorithm is Hamiltonian Monte Carlo (HMC) as implemented in the Stan programming language. The objective is to analyse the efficiency of the MCMC and their ability to return correct posterior estimates.

    Comparisons of these algorithms are conducted by Simulation Based Calibration \citep{talts2018validating}. Parameters are drawn from the prior distribution and used to create datasets from the generative stochastic volatility model which are sampled using one of the mentioned MCMC methods. Repeating this process multiple times gives insight to how well the MCMC algorithm can estimate the true parameters governing the data generating process. The key diagnostic metrics are the effective sample size to measure the efficiency of the algorithm and visual assessment of uniform rank statistics to determine calibration of the posterior estimates.

    Additionally, two different parametiersations of the stochastic volatility model are considered for each sampling method. An algorithm may be sensitive to different parameterisations of the same model which may impact MCMC performance as described in \citet{strickland2008parameterisation}. The key contribution of this research is to use simulations to assess whether a MCMC algorithm can return calibrated posterior estimates as well as the efficiency under the different parametiersations.

    This paper is structured as follows. Section 2 provides the context around this research, namely the stochastic volatility model and the limitations of single simulation studies and MCMC convergence diagnostics. Section 3 describes the simulation design and diagnostic metrics. Section 4 details the sampling approaches. Section 5 discusses the results and limitations. Finally, section 6 concludes and provides points for further reserach.

\section{Resarch Context}

\subsection{Stochastic Volatility}
    The model of interest is the discrete time, univariate stochastic volatility model estimated by \citet{kim1998stochastic} using Bayesian methods. $y_t$ is the mean corrected returns of some asset for equally spaced intervals t. $\beta$ is a constant scaling factor which is also defined as $exp(\mu / 2)$ representing instantaneous volatility. $h_t$ is log volatility, where $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the autoregressive parameter $\phi$ such that $|\phi|<1$. This autoregressive parameter represents the persistence or stickiness of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

    $$
    \begin{aligned}
    y_t =& \space \beta exp(h_t/2) \epsilon_t \\
    h_{t+1} =& \space \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t  \\
    h_1 \sim& \space normal\left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
    \end{aligned}
    $$

    $$
    \begin{aligned}
    \epsilon_t \sim& \space normal(0,1) \\
    \delta_t \sim& \space normal(0,1)
    \end{aligned}
    $$

    Setting $\beta=1$, the model can be expressed more succinctly as:

    $$
    \begin{aligned}
    y_t \sim& \space normal(0, exp(h_t/2)) \\ 
    h_1 \sim& \space normal \left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right) \\
    h_{t+1} \sim& \space normal(\mu +\phi(h_t - \mu) , \sigma_{\eta}^2), \space\space t\neq 1\\ 
    \end{aligned}
    $$

    Priors for the static parameters are defined below with conjugate priors on $\mu$ and $\sigma^2$:

    $$
    \begin{aligned}
    \mu \sim& \space normal(0, 10^2) \\
    \sigma_{\eta}^2 \sim& \space IG(5/2, (0.01\times 5) / 2) \\
    \phi^{\ast} \sim& \space beta(20, 1.5) \\
    \phi &=  2\phi^{\ast} - 1
    \end{aligned}
    $$

    The prior on $\phi$ is a stretched beta distribution. This is a beta distribution (as defined on the parameter $\phi^*$) which has been transformed to have support (-1, 1).

\subsection{Diagnostic limitations on real data}
    The stochastic volatility model can be estimated using a variety of sampling strategies and MCMC algorithms. Convergence metrics are often used to check the performance of the MCMC chains, such as the effective sample size and the $\hat{R}$. These diagnostics provide evidence on whether a MCMC has failed to converge onto the target posterior distribution.

    Convergence diagnostics are useful for identifying when Bayesian computation fail on real data. However, confounding issues may arise when attempting to diagnose the cause of computational problems. Real data is generated from an unknown data generating process. That is, the true parameter or model is unobservable. Therefore, failed diagnostic checks could arise from either inocrrect model specification or issues in the MCMC algorithm or both. Furthermore, different sampling strategies could provide different posterior estimates for the same model. Assuming no issues in computation, convergence diagnostics do not provide any information to which estimate is more correct or closer to the true parameter.

    To illustrate this problem, the stochastic volatility model is estimated with two different MCMC algorithms on real data. The data are the de-meaned daily log returns of the 2023 January-September S\&P 500 Index. Figure \ref{fig:realdataex} visualises the marginal posterior estimates for the static parameters $\mu$, $\phi$ and $\sigma^2$ with corresponding summary statistics in Table \ref{tab:realdata}.

    These estimates show the marginal posteriors of each parameter sharing similar shapes. $\phi$ has a fatter left tail for the KSC sampler and $\sigma^2$ is also heavier in the right tails relative to HMC. This is evident in the quantiles with the 25th quantile for $\phi$ being smaller at 0.875 for KSC compared to 0.910 in HMC. Furthermore, HMC has higher peaks at their mode relative to KSC with a tighter spread. Based on this information alone, there is no certainty in whether method 1 or method 2 is has a more correct estimate conditional on data and model.

@@

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{motivating_example/real_data_ex.png}
        \caption{Posterior samples from the Hamiltonian Monte Carlo (HMC) and Kim Sherphard and Chib (KSC) MCMC samplers using S\&P 500 data. While the shape of the distributions are similar, the modes of HMC is much higher and the tails of KSC are fatter for $\phi$ and $\sigma^2$. Convergence diagnostics do not tell us which estimate is more correct since we don't observe the true model or true data generating process.}
        \label{fig:realdataex}
    \end{figure}

    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|} \hline 
        Parameter&  MCMC&Min& q25&  Median& Mean & q75&Max\\ \hline 
        $\mu$&  KSC&-2.03 & -0.472 & -0.354 & -0.355 & -0.231 & 1.31 \\
     $\mu$&  HMC&-4.55 & -0.508 & -0.373 & -0.370 & -0.229 &2.86  \\\hline 
     $\phi$&  KSC&0.384 & 0.875 & 0.928 & 0.902 & 0.958 & 0.997 \\
     $\phi$&  HMC&0.438 & 0.910 & 0.953 & 0.929 & 0.977 &1.00  \\ \hline 
     $\sigma^2$&  KSC&0.00237 & 0.00891 & 0.0138 & 0.0178 & 0.0210 & 0.130 \\ 
     $\sigma^2$&  HMC&0.00209 & 0.00731 & 0.0105 & 0.0131 & 0.0159 &0.107 \\ \hline
        \end{tabular}
        \caption{Summary statistics for HMC and KSC algorithms. KSC's 25th quantile $\phi$ is smaller than HMC which is consistent with the fatter left tail. Similarly, the KSC's 75th qunatile for $\sigma^2$ is larger consistent with a larger right tail spread. Both sets of estimate are reasonable but it is unclear from estimates on real data which estimate is closer the the truth.}
        \label{tab:realdata}
    \end{table}

\subsection{Diagnostic limitations on a single simulation}
    Diagnosing computational problems on real data is difficult due to confounding factors. A strategy around this is to evaluate a model and algorithm on simulated data. One approach is to simulate data from a generative model using known parameters. Then fit the same model on the simulated data and see if the true parameters can be recovered. This gives us the benefit of defining the true parameters of the data generating process to be estimated. If the model and algorithm cannot adequately capture the true parameter, then we cannot be confident that it will provide reliable estimates on real data.

    Furthermore, as discussed in \citet{gelman2020bayesian}, fitting models on simulated data is the only way we can check inference on latent variables. This is critical in stochastic volatility since the underlying framework is a state space model with latent log volatility parameters. Latent variables are unobserved in real data and are only estimated in the context of the model. Simulation gives control over the data generating process which reveals what the model can infer about the latent variables.

    Simulations enable us to check whether our models and algorithms can estimate the true data generating process. However, there are limitations to what can be learned from a single simulation. There is always a small probability that the true parameter is in the tails of any estimated posterior distribution. Or put differently, there is a 1\% chance the true parameter or any random draw exists outside a 99\% credible interval. \cite{talts2018validating} make the point that a single simulation does not provide sufficient information about the inference made by an algorithm. As discussed in their paper, a single simulation may conclude ``that an incorrectly coded analysis worked as desired, while a correctly coded analysis failed''.

    Figure 2 show the posterior distributions of a single simulation where data is generated and posteriors are sampled from the stochastic volatility model with known parameters.

    \begin{figure}[h]
        \centering
        \includegraphics[scale=0.1]{motivating_example/single_sim.png}
        \caption{Posterior samples from known data genearting process. Vertical solid line represents true parameter and dotted lines are 95\% credible intervals around the mean. The parameter $\phi$ falls outside the credible interval whereas $\mu$ and $\sigma^2$ stay inside. In any one simulation conditional on the correct model, there is a 5\% probability that the true parameter falls outside this interval by chance.}
    \end{figure}

    The 95\% credible intervals for $\mu$ and $\sigma^2$ cover the true parameter. The true parameter for $\phi$ however, is in the tails and outside the interval. Such an analysis may incorrectly conclude that the model fails to adequatly estimate the $\phi$ parameter. However, it may be the case that the posterior distribution is correctly calculated using this algorithm and the results may be due to the features of this specific simulated dataset.

\subsection{Research Goal}
    The objective of this research is to design a simulation study to evaluate the calibration of algorithms used to estimate stochastic volatility models. As discussed, there are limitations to evaluating MCMC algorithms based on fits to real data and single simulations. To check the calibration of an algorithm, simulations over multiple true parameters are required. This methodology is discussed in the next section.

    Additionally, both the HMC and KSC algorithms will be assessed using the proposed simulation design as well as the performance of the samplers under different parameterisations of the stochastic volatility model. Results from the study will also be used to compare both sampling strategies to determine which MCMC approach is most suitable for estimating this model.

@@

    \subsection{Simulation Design}
        Simulation Based Calibration (SBC) checks the calibration of posterior estimates generated by MCMC algorithms. SBC is conducted by comparing the distribution of rank statistics to the discrete uniform distribution which arises when an algorithm is correctly calibrated. The procedure starts by taking draws from the prior distribution and creating datasets implied by each draw. Rank statistics are then calculated on the posterior samples conditional on the simulated data.

        To illustrate this procedure, let $\theta$ be an arbitrary parameter from a model and $y$ represent a dataset. Start with a single draw from the prior distribution:

        $$
        \begin{aligned}
        \theta^{sim} \sim \pi(\theta)
        \end{aligned}
        $$

        Generate a dataset given by the prior draw.

        $$
        \begin{aligned}
        y^{sim} \sim \pi (y|\theta^{sim})
        \end{aligned}
        $$

        Then take draws from the posterior distribution generated by a MCMC algorithm (HMC or KSC) conditional on this dataset.

        $$
        \begin{aligned}
        \{\theta_1,\dots , \theta_{L}\} \sim \pi (\theta | y^{sim})
        \end{aligned}
        $$

        A key result is that the posterior sample $\{\theta_1,\dots , \theta_{L}\}$ will share the same distribution as the prior samples $\theta^{sim}$. This is implied by the following expression:

        $$
        \begin{aligned}
        \pi(\theta) &= \int \pi(\theta|y^{sim}) \pi(y^{sim}|\theta^{sim}) \pi(\theta^{sim})dy^{sim} d\theta^{sim} \\
        &= \int \pi(\theta|y^{sim}) \pi(y^{sim},\theta^{sim}) dy^{sim} d\theta^{sim}
        \end{aligned}
        $$

        That is, the posterior averaged over the joint distribution follows the same distribution as the prior. The procedure of generating posterior samples implicitly performs this integral since the expression on the right of the integral is proportional to the prior density. Therefore, any deviation of the posterior distribution from the prior distribution means that the sampling methodology is not producing the correct posteriors.

        \citet{talts2018validating} prove that the rank statistics for a given parameter follows a discrete uniform distribution if the posterior samples follow the same distribution as the prior. The rank statistic is defined as:

        $$
        \begin{aligned}
        r = rank(\{\theta_1,\dots , \theta_{L}\}, \theta^{sim}) = \sum_{l=1}^{L}1[\theta_{l} < \theta^{sim}]
        \end{aligned}
        $$

        This completes one iteration of SBC. To complete the algorithm, multiple iterations are run and the rank statistics are calculated for each parameter. The resulting rank statistics are compared to the discrete uniform distribution to determine if the algorithm is calibrated and returning the correct posteriors.

        Posterior credible intervals are said to have sufficient coverage if Bayesian computation is well calibrated and the rank statistics follow a discrete uniform distribution. That said, there are two ways to interpret calibration.

        \begin{enumerate}
            \item For any percentage interval selected over the posterior samples (for example 90\%) then there is a 90\% chance that $\theta^{sim}$ falls within this interval.

            \item Bayesian analysis is well calibrated if a 90\% credible interval contains the true parameter in 90\% of the SBC iterations. 
        \end{enumerate}

@@

    \subsection{Evaluation Metrics}
        The key metrics to compare the performance of these methods are rank statistics and chi-squared test statistics to measure calibration, and the effective sample size (ESS) to measure effciency.

        \subsubsection{Rank statistics}
            Rank statistics are used to evaluate the calibration of the MCMC algorithm. If a posterior is well calibrated then it is expected that the distribution of rank statistics is uniform.

            The shape of the rank statistic distribution gives insight into how a MCMC may be miscalibrated. Specifically, it gives information about the type of bias in the posterior estimates for a given parameter. Figures \ref{fig:underestimation} and \ref{fig:underdispersed} are recreated from \citet{talts2018validating} for convenience which display examples of posterior estimates that give biased rank statistics.

            Figure \ref{fig:underestimation} shows a distribution of rank statistics with a large right peak. This is consistent with posterior samples underestimating the true parameter, where the estimated posterior distrubtion is biased to the left of the prior. The sum of indicator random variables is large since the majority of posterior draws are smaller than the true value. This results in a large rank statistic and a bias on the right side of the histogram. The coverse is true if the peak of the histogram is on the left hand side (i.e the posterior is biased towards the right hand side of the prior).

            Figure \ref{fig:underdispersed} has large peaks on both ends of the histogram. The sampler is disproportionately over and under estimating the true parameter. \citet{talts2018validating} describe this bias as underdispersion of the posterior relative to the prior distribuion. The estimated posterior is too narrow relative to the spread of the prior resulting in bias on both ends of the rank statistic distribution.

>>>>

\title{Comparing MCMC algorithms for Stochastic Volatility Model Bayesian Posteriors using Simulation Based Calibration}
\author{Benjamin Wee}
\date{October 2023}

@@

\begin{abstract}
    This research compares the computational methods used to estimate Bayesian stochastic volatility models. The MCMC approaches outlined in the stochastic volatility literature and Hamiltonian Monte Carlo are assessed in their ability to sample from the model's posterior distribution. Specifically, Simulation Based Calibration (SBC) is used to check whether these MCMC algorithms are returning efficient and calibrated posterior estimates. Key metrics of interest are the effective sample size to check the efficiency of the algorithm and tests of uniformity to assess the calibration of the posteriors. This will determine which method is better at estimating stochastic volatility models based on the efficiency and accuracy of the sampling strategy. Results reveal that Hamiltonian Monte Carlo provides more efficient and calibrated posterior estimates of the stochastic volatility model conditional on model parameterisation. 
\end{abstract}

@@

\section{Introduction}
    Stochastic volatility (SV) models are used in financial econometrics to model and predict the behaviour of financial asset returns. Typically expressed as a non-Gaussian state space model where the conditional variance of the observed return is treated as a latent random variable \citep{hull1987pricing, chesney1989pricing}, such models are difficult to estimate since the likelihood is unavailable in closed form. With more parameters than observations,\footnote{We refer to both static unknowns and time-varying latent state variables as 'parameters'.} SV models are commonly analysed using Bayesian methods obtained via a Markov Chain Monte Carlo (MCMC) algorithm designed to sample from the high dimensional joint posterior distribution. The focus of this paper is to investigate the performance of two competing MCMC algorithms, each applied to a SV model, in terms of the validity and statistical efficiency of the their resulting inference. Performance of these algorithms will also be compared with different parameterisations of the SV model to determine the sensitivity of each MCMC to model specification.

    \citet{kim1998stochastic} propose a sampling strategy for the so-called 'vanilla' stochastic volatility through an off-set mixture approximation. Conditional on the mixture allocations and static parameters, the forward filter backward sampling algorithm (Carter and Kohn, 1995; Fr{\"u}wirth-Schnatter, 1995) is used to jointly sample the latent states while conditionally  conjugate prior distributions are also suggested.  Since then, there have been many advances in statistical computing and algorithm design. In particular, Hamiltonian Monte Carlo (HMC) has be proposed as a generic and efficient sampling method for MCMC. Notably, HMC has been shown to be particularly well-suited to complex, high dimensional probabilistic models. Importantly, the adoption of such new techniques have been made widely available through the development of various open source libraries such as the Stan programming language \citep{stan} and the PyMC library \citep{pymc2023}. As the development of new algorithms rapidly increase, so does the need to develop new methods to assess their output. Developments in statistical workflow are required to test new algorithms as well as compare computational strategies used to estimate increasingly complex models.

    This research analyses and compares the calibration and efficiency of MCMC used to estimate stochastic volatility models using a simulation study. By calibration we refer to a MCMC algorithm that returns correct posterior, on average, conditional on the correct model specification. The first algorithm replicates Kim Shephard and Chib's (KSC) Gaussian mixture approximation which is sampled using the aforementioned Kalman Filter and Metropolis Hastings algorithm. The second algorithm is the No-U-turn sampler (NUTS) variant of Hamiltonian Monte Carlo (HMC) as implemented in the Stan programming language \citep{hoffman2014no, betancourt2017conceptual, stan}. The objective is to analyse the efficiency of the MCMC and their ability to return correct posterior estimates.

    Comparisons of these algorithms are conducted using a methodology known as Simulation Based Calibration (SBC) \citep{talts2018validating}. Parameters are drawn from the prior distribution and used to create data sets from the generative stochastic volatility model. We then obtain posterior samples from each of the MCMC methods. Repeating this process multiple times gives insight to how well the MCMC algorithm can estimate the true parameters given the data generating process. The key diagnostic metrics are the effective sample size, to measure the efficiency of the algorithm, and uniform rank statistics to determine calibration of the posterior estimates.

    One particular challenge to comparing models is the parameterisation. Different parameterisations of a model, whilst the same mathematically, can be very different in terms of sampling difficulty \citep{neal2003slice}. In this work two different parameterisations of the stochastic volatility model are compared for each sampling method. An algorithm may be sensitive to different parameterisations of the same model which may impact MCMC performance as described in \citet{strickland2008parameterisation}. The key findings of this research are HMC applied on a re-parameterised SV model produces the most calibrated and efficient posterior estimates and that the MCMC algorithms may perform better or worse based on calibration and efficiency depending on the parameterisation of the model.

    This paper is structured as follows. Section 2 provides the context around this research through the introduction of the stochastic volatility model and a discussion of the limitations of using ad-hoc simulations and MCMC convergence diagnostics to determine posterior calibration (or validity). Section 3 describes the SBC methodology, along with the simulation design and diagnostic metrics used for the SV model. Section 4 details key aspects of the two MCMC sampling approaches under consideration, while Section 5 discusses the results and limitations. Finally, Section 6 concludes and provides points for further research.

\section{Research Context}

\subsection{Stochastic Volatility}
    The model of interest is the discrete time, uni-variate stochastic volatility model estimated by \citet{kim1998stochastic} using Bayesian framework. SV is expressed as a state space model which links the observed mean corrected (logarithmic) return of a given asset to the latent conditional variance. 
    \begin{align}
    y_t =& \space \exp(h_t/2) \varepsilon_t, \\
    h_{t+1} =& \space \mu +\phi(h_t - \mu) + \sigma_{\eta} \eta_t,  \\
    h_1 \sim& \space \,\mathrm{N}\left(\mu, \frac{\sigma_{\eta}^2}{1-\phi^2}\right), \\
    \varepsilon_t \sim& \space \,\mathrm{N}(0,1), \\
    \eta_t \sim& \space \,\mathrm{N}(0,1).
    \end{align}
    $y_t$ is the mean corrected (logarithmic) return of a given asset, observed over an interval with fixed length and associated with period t. The measurement equation (1) describes the conditional distribution of $y_t$ given $h_t$, where $h_t$ is the conditional log variance of $y_t$. $h_1$ is a draw from a stationary distribution and the state equation $h_{t+1}$ follows a stationary process governed by the auto-regressive parameter $\phi$ such that $|\phi|<1$. This auto-regressive parameter represents the persistence or stickiness of log volatility and the dispersion parameter $\sigma_{\eta}$ is the constant variance of the states. $\epsilon_t$ and $\eta_t$ are standard normal white noise shocks and are uncorrelated with each other.

    KSC assign certain independent prior distributions and hyperparameters to the static parameters, with conjugate priors for $\mu$ and $\sigma^2$, given by
    \begin{align}
    \mu \sim& \space \,\mathrm{N}(0, 10^2) \\
    \sigma_{\eta}^2 \sim& \space \, \mathrm{IG}(shape=5/2, scale=(0.01\times 5) / 2) \\
    \phi^{\ast} \sim& \space \,\mathrm{Beta}(20, 1.5) \\
    \phi =&  2\phi^{\ast} - 1
    \end{align}
    The prior distribution for $\phi$ is referred to as a stretched beta distribution, as it corresponds to a transformed beta random variable, $\phi^*$, with support (-1, 1).

\subsection{Challenge 1: Comparing posteriors with no ground truth}
    The stochastic volatility model can be estimated using a variety of sampling strategies and MCMC algorithms. Convergence metrics assessed from the same MCMC output to be used for posterior inference are commonly used to check the performance of the MCMC chains. Most often, the effective sample for each parameter is produced, as it indicates the number of approximately independent draws obtained from the dependent Markov chain for a target parameter \citep{gelman2013bayesian}, and $\hat{R}$, the potential scale reduction factor or measure of variability between different chains \citep{gelman1992inference}, is computed, to gauge whether the MCMC algorithm has converged to the target posterior distribution.

    Convergence diagnostics are useful for identifying when Bayesian computation fail on real data. However, confounding issues may arise when attempting to diagnose the cause of computational problems. Real data is generated from an unknown data generating process. That is, the true parameter and model are unobservable, and it is very likely that the model being fit is not the true model. Therefore, failed diagnostic checks could be the result of attempting to fit a misspecified model, issues with the MCMC algorithm, or both. Furthermore, different sampling strategies will result in different posterior estimates for the same model. Assuming no issues with the computation or the model specification, convergence diagnostics do not provide any guidance to the analyst regarding which posterior estimate is to be preferred since there is no ground truth.

    To illustrate the concerning issues, the posterior distribution associated with the stochastic volatility model in (1)-(5) and a set of real financial returns is estimated using two different MCMC algorithms. The data are demeaned daily (logarithmic) returns on the S\&P 500 Index over the period from 1 January 2023 - 10 September 2023 with sample size $T=171$. Figure \ref{fig:realdataex} displays the marginal posterior estimates for the static parameters $\mu$, $\phi$ and $\sigma^2$, with corresponding summary statistics in Table \ref{tab:realdata}. Overall the two methods are similar in skew and shape. However, closer inspections suggests that $\phi$ has a fatter left tail for the KSC sampler and $\sigma^2$ is also heavier in the right tails relative to HMC. This is evident in the quantiles with the 25th quantile for $\phi$ being smaller at 0.875 for KSC compared to 0.910 in HMC. Furthermore, HMC has taller peaks at their mode relative to KSC with a tighter spread. The uncertainty intervals created from each posterior sample would also be quite different and lead to different results. However, we can only conclude that there are differences between these methods, not which is correct.

@@

    \begin{figure}[H]
        \centering
        \includegraphics[scale=0.1]{motivating_example/real_data_ex.png}
        \caption{\textbf{Posterior distribution for static parameters fit on the S\&P 500 Index.} Posterior samples from the HMC (blue) and KSC (red) MCMC samplers.}
        \label{fig:realdataex}
    \end{figure}

    \begin{table}[H]
        \centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|} \hline 
        Parameter&  MCMC&Min& q25&  Median& Mean & q75&Max\\ \hline 
        $\mu$&  KSC&-2.03 & -0.472 & -0.354 & -0.355 & -0.231 & 1.31 \\
     $\mu$&  HMC&-4.55 & -0.508 & -0.373 & -0.370 & -0.229 &2.86  \\\hline 
     $\phi$&  KSC&0.384 & 0.875 & 0.928 & 0.902 & 0.958 & 0.997 \\
     $\phi$&  HMC&0.438 & 0.910 & 0.953 & 0.929 & 0.977 &1.00  \\ \hline 
     $\sigma^2$&  KSC&0.00237 & 0.00891 & 0.0138 & 0.0178 & 0.0210 & 0.130 \\ 
     $\sigma^2$&  HMC&0.00209 & 0.00731 & 0.0105 & 0.0131 & 0.0159 &0.107 \\ \hline
        \end{tabular}
        \caption{\textbf{Summary statistics for HMC and KSC algorithms}. Both sets of estimates appear reasonable but it is unclear from estimates on real data which estimate is closer the the truth.}
        \label{tab:realdata}
    \end{table}

\subsection{Challenge 2: Limitations of a single simulation}
    Diagnosing problems with the approximate posterior from models fit to real data is difficult because we do not have a ground truth to compare with. A strategy around this is to evaluate a model and algorithm on simulated data. One approach is to simulate data from a generative model using known parameters. Then fit the same model on the simulated data and see if the true parameters can be recovered. This gives us the benefit of defining the true parameters of the data generating process to be estimated. If the model and algorithm cannot adequately capture the true parameter, then we cannot be confident that it will provide reliable estimates on real data.

    Simulation is particularly important in the SV model as it has latent parameters. \citet{gelman2020bayesian} discuss that simulation is the only way we can check inference on latent variables. This is critical for the SV model since the underlying framework is a state space model with latent log volatility parameters. Latent variables are unobserved in real data and are only estimated in the context of the model. Simulation gives control over the data generating process which reveals what the model can infer about the latent variables.

    Simulations enable the analyst to check whether a model and corresponding simulation algorithms appropriately estimate the `true' data generating process. However, there are limitations to what can be learned from a single simulation. Even for a correctly specified model, there is always a small probability that the true parameter is in the tails of the corresponding (exact) marginal posterior distribution. \cite{talts2018validating} make the point that a single simulation does not provide sufficient information about the inference made by an algorithm. As discussed in their paper, a single simulation may conclude ``that an incorrectly coded analysis worked as desired, while a correctly coded analysis failed''.

    Figure 2 shows marginal posterior distributions given a single set of 1000 returns generated by the SV model. The true parameters are $\mu=-0.389$, $\sigma^2_{\nu}=0.0115$ and $\phi=0.547$ and the model is fit using Stan's implementation of HMC, the No-U-Turn sampler. The 95\% credible intervals for $\mu$ and $\sigma^2$ cover the true parameter. The true parameter for $\phi$ however, is in the tails and outside the interval. Such an analysis may incorrectly conclude that the model fails to adequately estimate the $\phi$ parameter. However, it may be the case that the posterior distribution is correctly calculated using this algorithm and the results may be due to the features of this specific simulated data set.

    \begin{figure}[h]
        \centering
        \includegraphics[scale=0.1]{motivating_example/single_sim.png}
        \caption{\textbf{Posterior samples from SV model fit on known data generating process}. Vertical solid line represents true parameter and dotted lines are 95\% credible intervals around the mean. The parameter $\phi$ falls outside the credible interval whereas $\mu$ and $\sigma^2$ stay inside.}
    \end{figure}

\subsection{Research Goal}
    The objective of this research is to design a simulation study to evaluate the calibration of the HMC and KSC offset-mixture algorithms used to estimate SV models. As discussed, there are limitations to evaluating MCMC algorithms based on fits to real data and single simulations. To check the calibration of an algorithm, repeated independent simulations are required. This methodology is discussed in the next section.

    Two parameterisations of the SV model for each sampler will be implemented using this simulation design. Results from the study will also be used to compare both sampling strategies to determine which MCMC approach is most suitable for estimating this model.

@@

    \subsection{Simulation Design}
        Simulation Based Calibration (SBC) checks the calibration of posterior estimates generated by MCMC algorithms. SBC is conducted by comparing the distribution of rank statistics to the uniform distribution which arises when an algorithm is correctly calibrated. For each $k$ of $K$ SBC iterations, one value for each parameter is drawn from the prior. For this parameter a data set is generated from the model. Then a sampling method (originally using Stan in \citet{talts2018validating}) is used to obtain posterior samples given this data and the prior. Then the estimated posteriors are compared to the true values by calculating the rank. Under a calibrated model, the $K$ rank statistics for each parameter should be uniformly distributed.

        To illustrate the SBC procedure, let $\theta$ be an arbitrary scalar parameter from a model and $y$ represent a data set. Start with a single parameter $\theta^{sim}$ drawn from the prior distribution:
        \begin{align}
        \theta^{sim} \sim \pi(\theta),
        \end{align}
        generate a data set $y^{sim}$ conditionally given $\theta^{sim}$ and the model of interest\footnote{Since all computation presumes a correctly specified model, nottation to condition on the model is suppressed.}:
        \begin{align}
        y^{sim} \sim \pi (y|\theta^{sim}).
        \end{align}
        Then take $L$ draws from the estimated posterior distribution, conditional on this data set, and generated by the MCMC algorithm under investigation (here, HMC or KSC):
        \begin{align}
        \{\theta_1,\dots , \theta_{B}\} \sim \pi (\theta | y^{sim}).
        \end{align}

        \citet{talts2018validating} prove that the sample of rank statistics for a given parameter, produced from replicate posteriors, follow a uniform distribution if the posterior samples follow the same distribution as the prior on average. The rank statistic for $\theta$ produced by the $k^{\mathrm{th}}$ replicate posterior sample, is defined as
        \begin{align}
        r^{(k)} = rank(\{\theta_1^{(k)},\dots , \theta_{L}^{(k)}\}, \theta^{sim}) = \sum_{b=1}^{B}1[\theta_{b}^{(k)} < \theta^{sim}].
        \end{align}

        This completes one iteration of SBC. To complete the algorithm, multiple iterations are run and the rank statistics are calculated for each parameter. The resulting rank statistics are compared to the uniform distribution to determine if the algorithm is calibrated and returning the correct posteriors.

        A key result is that the posterior averaged over the data and true parameters will equal the prior distribution. This is shown in the following expression:
        \begin{align}
        \pi(\theta) &= \int \int \pi(\theta|y^{sim}) \pi(y^{sim}|\theta^{sim}) \pi(\theta^{sim})d\theta^{sim} dy^{sim}, \\
        &= \int \int \pi(\theta|y^{sim}) \pi(y^{sim},\theta^{sim}) d\theta^{sim} dy^{sim}.
        \end{align}
        That is, the average posterior for $\theta$ over the SBC iterations should approximately equal to prior distribution, if the algorithm is calibrated. Therefore, any deviation from the prior distribution and the average posterior draws produced from replicated posterior distributions means that the sampling methodology is not producing the correct posterior, on average.

        Posterior credible intervals are said to have sufficient coverage if Bayesian computation is well calibrated and the rank statistics follow a uniform distribution. That is, one way to describe calibration is: for any percentage interval selected over the posterior samples (for example 90\%) then there is a 90\% chance that $\theta^{sim}$ falls within this interval \citep{dawid1982well}.

@@

    \subsection{Evaluation Metrics}
        In this research the following metrics are chosen:

        \begin{enumerate}
            \item Rank statistics to measure calibration for a given parameter and algorithm.

            \item Chi squared statistics calculated over the $K$ rank statistics to summarise the shape of the distribution and deviation from uniformity.

            \item Effective sample size (ESS) to measure MCMC efficiency
        \end{enumerate}

        \subsubsection{Rank statistics}
            Rank statistics are used to evaluate the calibration of the MCMC algorithm. If a posterior is well calibrated then it is expected that the distribution of rank statistics over $B$ SBC iterations for each parameter in the model is uniform.

            The shape of the distribution of rank statistics gives insight into how an MCMC may be miscalibrated. Specifically, it gives information about the type of bias in a marginal posterior estimate of a given parameter. Figures \ref{fig:underestimation} and \ref{fig:underdispersed} are recreated from \citet{talts2018validating} for convenience. In these we display examples of posterior sampling bias and the corresponding non-uniform rank statistics.

            Figure \ref{fig:underestimation} shows a distribution of rank statistics with a large right peak. This distribution is consistent with posterior samples that tend to under estimate the true parameter on average. The sum of indicator random variables (see equation (13)) is large since a large proportion of posterior draws are smaller than the true value, resulting in a large rank statistic and hence a larger relative frequency on the right side of the histogram. The converse is true if the peak of the histogram is on the left hand side (i.e the average posterior over the SBC iterations is biased towards the right hand side of the prior).

            Figure \ref{fig:underdispersed} has large peaks on both ends of the histogram. \citet{talts2018validating} describe this bias as under-dispersion of the posterior relative to the prior distribution on average. The estimated posterior is too narrow relative to the spread of the prior resulting in oversampling on both ends of the rank statistic distribution.
